---
title: "ml_proj_report"
output: html_document
---
# Summary

# Prep
Split into training and test datasets


## Exploratory
### create boxplots

### Look for excessive NAs

### Variables that are highly correlated with each other

### Train model

#### Discussion of model output
Accuracy 99%
Cross-validation - use default bootstrap with 25 repetitions
Number of Trees: 500
OOB estimate of error: ~0.6%

```
> dim(train)
[1] 13737    28

> dim(test)
[1] 5885   20
```


The Final Model
```{r, eval=FALSE}
modFit
```
```
Random Forest 

13737 samples
   27 predictors
    5 classes: 'A', 'B', 'C', 'D', 'E' 

No pre-processing
Resampling: Bootstrapped (25 reps) 

Summary of sample sizes: 13737, 13737, 13737, 13737, 13737, 13737, ... 

Resampling results across tuning parameters:

  mtry  Accuracy  Kappa  Accuracy SD  Kappa SD
  2     0.99      0.988  0.0015       0.0019  
  14    0.989     0.986  0.00178      0.00225 
  27    0.984     0.98   0.00285      0.0036  

Accuracy was used to select the optimal model using  the largest value.
The final value used for the model was mtry = 2. 
```

```{r, eval=FALSE}
modFit$finalModel
```
```
Call:
 randomForest(x = x, y = y, mtry = param$mtry, proximity = TRUE) 
               Type of random forest: classification
                     Number of trees: 500
No. of variables tried at each split: 2

        OOB estimate of  error rate: 0.6%
Confusion matrix:
     A    B    C    D    E  class.error
A 3903    1    0    0    2 0.0007680492
B   15 2630   13    0    0 0.0105342363
C    0   10 2373   13    0 0.0095993322
D    0    0   20 2230    2 0.0097690941
E    0    0    1    5 2519 0.0023762376
```



Confusinon matrix
```{r, eval=FALSE}
confusionMatrix(pred, test$classe)
```
```
Confusion Matrix and Statistics

          Reference
Prediction    A    B    C    D    E
         A 1672    7    0    0    0
         B    1 1129    3    0    0
         C    0    3 1015   13    1
         D    0    0    8  951    4
         E    1    0    0    0 1077

Overall Statistics
                                         
               Accuracy : 0.993          
                 95% CI : (0.9906, 0.995)
    No Information Rate : 0.2845         
    P-Value [Acc > NIR] : < 2.2e-16      
                                         
                  Kappa : 0.9912         
 Mcnemar's Test P-Value : NA             

Statistics by Class:

                     Class: A Class: B Class: C Class: D Class: E
Sensitivity            0.9988   0.9912   0.9893   0.9865   0.9954
Specificity            0.9983   0.9992   0.9965   0.9976   0.9998
Pos Pred Value         0.9958   0.9965   0.9835   0.9875   0.9991
Neg Pred Value         0.9995   0.9979   0.9977   0.9974   0.9990
Prevalence             0.2845   0.1935   0.1743   0.1638   0.1839
Detection Rate         0.2841   0.1918   0.1725   0.1616   0.1830
Detection Prevalence   0.2853   0.1925   0.1754   0.1636   0.1832
Balanced Accuracy      0.9986   0.9952   0.9929   0.9920   0.9976
```

### Most important variables
```{r, eval=FALSE}
varImp(modFit)
```
```
rf variable importance

  only 20 most important variables shown (out of 27)

                  Overall
roll_belt          100.00
yaw_belt            78.91
magnet_dumbbell_z   61.87
pitch_forearm       61.70
pitch_belt          55.43
roll_dumbbell       44.11
roll_forearm        42.63
roll_arm            31.21
gyros_belt_z        29.18
yaw_dumbbell        27.41
magnet_forearm_z    27.14
gyros_dumbbell_y    26.93
yaw_arm             26.12
pitch_dumbbell      24.99
magnet_forearm_y    23.17
yaw_forearm         18.98
pitch_arm           13.63
gyros_arm_y         11.72
gyros_dumbbell_x    11.17
gyros_arm_x         10.92
```