High Impact Weather Events
========================================================

## Synopsis



## Data Processing
Read in the b2zip Data
```{r}
# Abount 410 Mb in size once read in, 902297 rows
# zz <- bfile("StormData.csv.bz2", "r")
# data <- read.csv("StormData.csv", nrows=903000)
```

Summarize data along the lines
* Affects on humnan health - FATALITIES and INJURIES  for each EVTYPE
* Economic effects - PROPDMG, CROPDMG for each EVTYPE, look at PROPDMGEXP and CROPDMGEXP which look like give units (ex. K)
* Mean damage by type
* Look at linear model (scatterplot?)

```{r}
# Create smalle r data frame and release rest of original data
df2 <- subset(data, select=c(BGN_DATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP,CROPDMG,CROPDMGEXP))
rm(data)
df2$date <- as.Date(df2$BGN_DATE,"%m/%d/%Y")
df2$year <- as.character(df2$date, format="%Y")
# Create column for year and populte based on BGN_DATE
library(plyr)
# Need to normalize damage amount
yearly.prop.dmg <- ddply(df2, .(year, EVTYPE), summarize, propertydamage=sum(PROPDMG))
yearly.rollup <- ddply(yearly.prop.dmg, .(year), summarize, propdmg=sum(propertydamage))
```

Graphs to look at:
* Across time
* By event type
* Subsetting of health and damage components
* By geo area (time permitting)
* By month (time permitting)

## Results

You can also embed plots, for example:


```{r fig.width=7, fig.height=6}
#plot(cars)
```

