Most Harmful and Damaging Weather Event Types in the United States
========================================================

## Synopsis
In this report we aim to show what type of weather events have been the most harmful to human health and which event types have been the most destructive in terms of damage. Having this information can help with the prioritization of the limited government resources by helping focus on those event types which have the most impact. To investigate this we obtained the National Weather Service storm data, specifically for the years 1952 to 2011 (the most recent complete year available). From these data we found that flooding and tropical cyclones (hurricanes and tropical storms) have caused much more damage than other types of severe weather. For injuries and fatalities, no two events dominated to the degree like flooding and tropical systems did with damages, however, tornadoes and heat releated events caused had the most impact. Tornadoes caused more than forty percent of all injuries and heat releated events caused more than forty percent of fatalities.


## Loading and Processing the Raw Data
From the [Coursera Website](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) (49MB) we obtained the storm data. The single file contains a little over 900,000 observations from the year 1952 to 2011. Per the [national weather site](http://www.ncdc.noaa.gov/stormevents/details.jsp) only limited data was collected before 1995. Prior to 1995 only data for tornadoes, thunderstorm winds, and hail was collected.

### Reading in the data
First we read in the data contained in the bz2 archive. The data is delimeted with commas with some factors having leading and trailing spaces. There are 902297 rows, so we will specify the number of rows just a little higher.
```{r, cache=TRUE}
if (file.exists("Stormdata.csv")) {
    df2 <- read.csv("StormData.csv", nrows=903000, strip.white=TRUE)
} else if (file.exists("StormData.csv.bz2")) {
    df2 <- read.csv(bzfile("StormData.csv.bz2"),nrows=903000, strip.white=TRUE)
} else {
    stop("File not Found: Please download the StormData.csv.bz2 file.")
}
```

After reading in the data we check the first few rows (there are 902,297).
```{r}
dim(df2)
head(df2,5)
```

### Only keep variables we are interested in
We then only select the variables of interest for this analysis and remove the rest. This will give back memory.
```{r, cache=TRUE}
# Create smaller data frame and release rest of original data
df2 <- subset(df2, select=c(BGN_DATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP,CROPDMG,CROPDMGEXP,F,MAG))
gc()
```


### Limit data set to post-1994
Prior to 1995 only data for tornadoes, thunderstorm winds, and hail were gathered, which could skew results since those three categeries have an additional forty-two years of data([Limited collection before 1995](http://www.ncdc.noaa.gov/stormevents/details.jsp) ). We will keep only the data after 1994.

First we extract the year from the begin date in the data.
```{r, cache=TRUE}
df2$date <- as.Date(df2$BGN_DATE,"%m/%d/%Y")
df2$year <- as.character(df2$date, format="%Y")
```

Next we select only the post 1994 data
```{r, cache=TRUE}
df2 <- subset(df2, year > 1994)
```

### Put damage amounts into a common unit and remove leading spaces
Since the damage amounts are in different units, we create a function that will calculate the damage for a given base and exponent. In additions to powers of ten the characters H, K, M, and B (both upper and lower case) are used to indicates hundreds, thousands, millions, and billions respectively.
```{r}
dmgcalc <- function(base, exp) {
    ifelse(exp == "H" | exp == "h", base * 100,
    ifelse(exp == "K" | exp == "k", base * 1000,
    ifelse(exp == "M" | exp == "m", base * 1000000,
    ifelse(exp == "B" | exp == "b", base * 1000000000,
    ifelse(grepl("[0,9]*[.]*[0-9]",exp), base * 10^as.numeric(exp), 0
    )))))
}
```

For each record we will convert the property damage and the crop damage into dollars. While doing that we will also calculate the total damage and total health impact. Total damage here consists of the sum of the property damage and crop damage from a given event. We use this definition since both represent economic impact and both can occur from a single event. For health impact we use the sum of the reported fatalities and the reported injuries. Although most people would agree that a fatality has more impact than an injury, here we give them equal weight.  

We will clean up the EVTPE variable at the same time. This factor has leading spaces for some levels which causes them to be treated as a different instance in many functions. For example, "   HIGH TIDE" gets treated differently that "HIGH TIDE". We remove all leading and trailing spaces and then recreate the factor.
```{r, cache=FALSE}
library(stringr)
df2 <- within(df2, {
    PROPDMGEXP <- as.character(PROPDMGEXP)
    CROPDMGEXP <- as.character(CROPDMGEXP)
#    date <- as.Date(BGN_DATE,"%m/%d/%Y")
#    year <- as.character(date, format="%Y")
    proptot <- dmgcalc(PROPDMG, PROPDMGEXP)
    croptot <- dmgcalc(CROPDMG, CROPDMGEXP)
    totaldmg <- proptot + croptot
    healthcomp <- INJURIES + FATALITIES
    EVTYPE <- str_trim(as.character(EVTYPE))
    EVTYPE <- as.factor(EVTYPE)
})
```

We look at the first few records to verify conversion.
```{r}
head(df2)
```

Check for NAs being introduced into totals
```{r}
sum(is.na(df2$proptot))
sum(is.na(df2$croptot))
sum(is.na(df2$totaldmg))
sum(is.na(df2$healthcomp))
```


### Combine like event types
The data has many different strings to identify the same event type, for example "THUNDERSTORM" and "TSTM" for thunderstorms. Additionally, for some events the event is identified by name and not type such as "HURRICANE OPAL" instead of just "HURRICANE". We create a function to replace what we have identified as like type events with the same event type.
```{r, cache=TRUE}
cleantypes <- function(x) {
    # Change to a character so can more easily change factor levels
    x <- as.character(x)
    
    # Thunderstorms and Tornadoes
    x[grepl("TORNADO",x)] <- "TORNADO"
    x[grepl("TORNDAO",x)] <- "TORNADO"
    x[grepl("FUNNEL CLOUD",x)] <- "TORNADO"
    x[grepl("HAIL",x)] <- "HAIL"
    x[grepl("THUNDERSTORM",x)] <- "THUNDERSTORM"
    x[grepl("TSTM",x)] <- "THUNDERSTORM"
    x[grepl("Tstm",x)] <- "THUNDERSTORM"
    x[grepl("MICROBURST",x)] <- "MICROBURST"
    x[grepl("Microburst",x)] <- "MICROBURST"
    
    # Tropical Systems
    x[grepl("HURRICANE",x)] <- "HURRICANE/TROPICAL STORM"
    x[grepl("Hurricane",x)] <- "HURRICANE/TROPICAL STORM"
    x[grepl("TYPHOON",x)] <- "HURRICANE/TROPICAL STORM"
    x[grepl("TROPICAL STORM",x)] <- "HURRICANE/TROPICAL STORM"
    x[grepl("COASTAL",x)] <- "COASTAL STORM"
    x[grepl("Coastal",x)] <- "COASTAL STORM"
    x[grepl("SURGE",x)] <- "HURRICANE/TROPICAL STORM"
    
    # Heat Waves
    x[grepl("HEAT",x)] <- "HEAT"
    x[grepl("Heat",x)] <- "HEAT"
    x[grepl("WARM",x)] <- "HEAT"
    
    # High Winds
    x[grepl("WIND",x)] <- "HIGH WINDS"
    x[grepl("Wind",x)] <- "HIGH WINDS"
    x[grepl("wind",x)] <- "HIGH WINDS"
    
    # Excessive Rain
    x[grepl("RAIN",x)] <- "RAIN"
    x[grepl("Rain",x)] <- "RAIN"
    # Period where not enough time to dry out before next rain
    x[grepl("EXCESSIVE WETNESS",x)] <- "RAIN" 
    
    # Extreme Cold and Winter Storms
    x[grepl("WINTER",x)] <- "WINTER STORM"
    x[grepl("Winter",x)] <- "WINTER STORM"
    x[grepl("SNOW",x)] <- "WINTER STORM"
    x[grepl("Snow",x)] <- "WINTER STORM"
    x[grepl("BLIZZARD",x)] <- "WINTER STORM"
    x[grepl("WINTRY MIX",x)] <- "WINTER STORM"
    x[grepl("COLD",x)] <- "EXTREME COLD"
    x[grepl("cold",x)] <- "EXTREME COLD"
    x[grepl("Cold",x)] <- "EXTREME COLD"
    x[grepl("FREEZE",x)] <- "EXTREME COLD"
    x[grepl("LOW TEMP",x)] <- "EXTREME COLD"
    x[grepl("MIX",x)] <- "WINTERY MIX"
    x[grepl("Mix",x)] <- "WINTERY MIX"
    x[grepl("FREEZING",x)] <- "WINTERY MIX"
    x[grepl("Freezing",x)] <- "WINTERY MIX"
    
    # Coastal Surf and High Seas
    x[grepl("SURF",x)] <- "HIGH SURF/RIP CURRENTS"
    x[grepl("Surf",x)] <- "HIGH SURF/RIP CURRENTS"
    x[grepl("HIGH SEAS",x)] <- "HIGH SURF/RIP CURRENTS"
    x[grepl("RIP CURRENT",x)] <- "HIGH SURF/RIP CURRENTS"
    x[grepl("HIGH SWELLS",x)] <- "ROUGH SEAS"
    
    # Flooding
    x[grepl("FLASH FLOOD",x)] <- "FLASH FLOOD"
    x[grepl("FLOOD/FLASH",x)] <- "FLASH FLOOD"
    x[grepl("STREAM",x)] <- "FLASH FLOOD"
    x[grepl("FLOOD",x)] <- "FLOODING"
    
    # Forest and Wildfires
    x[grepl("FOREST FIRE",x)] <- "FOREST FIRES"
    x[grepl("WILD",x)] <- "WILDFIRES"
    
    # Lightning
    x[grepl("LIGHTNING",x)] <- "LIGHTNING"
    x[grepl("LIGHNTING",x)] <- "LIGHTNING"
    x[grepl("LIGNTNING",x)] <- "LIGHTNING"
    x[grepl("LIGHTING",x)] <- "LIGHTNING"
    
    # Miscellaneous
    x[grepl("MUD",x)] <- "MUD SLIDES"
    x[grepl("Mud",x)] <- "MUD SLIDES"
    x[grepl("ICE JAM",x)] <- "JAM"
    x[grepl("Ice jam",x)] <- "JAM"
    x[grepl("ICE FLOW",x)] <- "JAM"
    x[grepl("AVALANCE",x)] <- "AVALANCHE"
    x[grepl("HYPOTHERM",x)] <- "HYPOTHERMIA"
    x[grepl("HYPERTHERM",x)] <- "HYPOTHERMIA"
    x[grepl("Hypotherm",x)] <- "HYPOTHERMIA"
    x[grepl("LANDSLIDE",x)] <- "LANDSLIDE"
    x[grepl("MARINE",x)] <- "MARINE MISHAP"
    x[grepl("Marine",x)] <- "MARINE MISHAP"
    
    # Snow and Ice not able to directly associate with a storm
    x[grepl("ICE",x)] <- "SNOW/ICE"
    x[grepl("Ice",x)] <- "SNOW/ICE"
    x[grepl("GLAZE",x)] <- "SNOW/ICE"
    x[grepl("Glaze",x)] <- "SNOW/ICE"
    x[grepl("snow",x)] <- "SNOW/ICE"
    x[grepl("ICY",x)] <- "SNOW/ICE"
    x[grepl("SLEET",x)] <- "SNOW/ICE"
    
    # Change back into a factor
    x <- factor(x)
    x
}
```

We then use the function to create a new variable with this cleaned up set of event types.
```{r}
df2$cleanedEVTYPE <- cleantypes(df2$EVTYPE)
```


## Results
In order to show what type of events cause the most damage and harm we need to summarize by type of event.

### Create summary values
We summarize the data and compute the following for each event type:
- Total property damage
- Total crop damage
- Total of crop and property damage combined
- Average amount of property damage
- Average amount of crop damage
- Average amount of total damage
- Total number of fatalities
- Total number of injuries
- Total number of fatalities and injries compbined
- Average number of fatalities
- Average number of injuries
- Average amount of fatalities and injuries combined
```{r, cache=TRUE}
library(plyr)
# yearly.prop.dmg <- ddply(df2, .(year, cleanedEVTYPE), summarize, propertydamage=sum(totaldmg))
# yearly.rollup <- ddply(yearly.prop.dmg, .(year), summarize, propdmg=sum(propertydamage))
# yearly.dmg <- ddply(df2, .(year, cleanedEVTYPE), summarize, propertydamage=sum(totaldmg))
typesummary <- ddply(df2, ~ cleanedEVTYPE, summarize, prop=sum(proptot), avgprop=mean(proptot), 
                     crop=sum(croptot), avgcrop=mean(croptot),
                     damage=sum(totaldmg), avgdamage=mean(totaldmg),
                     fatal=sum(FATALITIES), avgfatal=round(mean(FATALITIES),0), 
                     injury=sum(INJURIES), avginjury=round(mean(INJURIES),0),
                     health=sum(healthcomp), avghealth=round(mean(healthcomp),0)
                     ) 
# yearly.dmg <- ddply(df2, .(year, cleanedEVTYPE), summarize, propertydamage=sum(totaldmg))
```

To simplify things we will remove all summary event types which have not damage or health impact.
```{r}
typesummary <- subset(typesummary, damage > 0 | health > 0)
typesummary <- droplevels(typesummary)
```

Since we are interested in the "most" impactful weather events we will create a function that will return the top N events of a passed data frame given a specific column to use for determining the largest. This way we can call the function with our data frame and just give a different column to find the top N events of that type based upon one of our calculated values.
```{r}
topvals <- function(df, col, num) {
    if (is.character(col)) {
        colnum <- which (colnames(df) == col)
    } else {
        colnum <- col
    }
    
    top <- with(df, df[order(-df[colnum])[1:num],])
    top
}
```

### Five weather event types that have caused the most damage
We will look at the five top event types for causing the most total economic damage over time. First we get the top five.
```{r}
btot <- topvals(typesummary, "damage", 5)
```

Now we take a look at them in a bar chart
```{r}
library(ggplot2)
ggplot(btot, aes(x=reorder(cleanedEVTYPE, order(damage, decreasing=TRUE)), 
                 y=damage/1000000000)) + geom_bar(stat="identity", fill="blue") + 
    xlab("Event Type") + ylab("Total Damage (billions)")
```

As can be seen, the top two type of events in terms of total damage, flooding and hurricanes/tropical storms, have caused many times more damage than any of the other types of events. 

### Five weather event types with most impact on health
Next we look at the five types of severe weather events that have had the largest impact on health. We will look at the five top event types for causing the most total economic damage over time. First we get the top five.
```{r}
health <- topvals(typesummary, "health", 5)
```

Now we take a look at them in a bar chart.
```{r}
ggplot(health, aes(x=reorder(cleanedEVTYPE, order(health, decreasing=TRUE)), 
                 y=health)) + geom_bar(stat="identity", fill="blue") + 
    xlab("Event Type") + ylab("Total of Injuries and Fatalities")
```

With injuries and fatalities no one or two event types dominates to the degree that we saw with damages. However, tornadoes still seem to have a substantially greater impact on injuries and fatalities than other types of severe weather. 

Since tornadoes have caused more injuries and death, lets see how tornadoes compare to all other weather events combined. For tornadoes we have:
```{r}
health[health$cleanedEVTYPE == "TORNADO", c("health","injury","fatal")]
```

and for all others we have:
```{r}
colSums(health[health$cleanedEVTYPE != "TORNADO",c("health","injury","fatal")])
```

Percentage-wise we have:
```{r}
library(scales)
ratios <- percent(colSums(health[health$cleanedEVTYPE == "TORNADO",c("health","injury","fatal")]) / colSums(health[,c("health","injury","fatal")]))
names(ratios) <- c("health","injury","fatal")
ratios
```

Tornadoes have caused more than forty percent of all injuries but only twenty-one  percent of all fatalities. Although twenty-one percent is a large number, will look to see if there is another severe weather type that has has caused a large number of fatalities.
```{r}
bfatal <- topvals(typesummary, "fatal", 5)
ggplot(bfatal, aes(x=reorder(cleanedEVTYPE, order(fatal, decreasing=TRUE)), 
                 y=fatal)) + geom_bar(stat="identity", fill="blue") + 
    xlab("Event Type") + ylab("Total Fatalities")
avghealth <- topvals(typesummary, "avghealth",5)
```

We see historically heat related events have caused the greatest number of fatalities.

```{r}
ratios <- percent(colSums(bfatal[bfatal$cleanedEVTYPE == "HEAT",c("health","injury","fatal")]) / colSums(bfatal[,c("health","injury","fatal")]))
names(ratios) <- c("health","injury","fatal")
ratios
```

We see that heat releated events have caused forty-one percent of fatalities due to severe weather. Tornadoes and heat events together have caused sixty percent of all injuries and deaths due to severe weather.
